Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-04 opencl/papers/foropencl.tex
more text, more refs

\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}
\author{Awesome Fortran Programmers}

@@

\begin{abstract}
I am abstract.
\end{abstract}

@@

This paper presents a compiler-level approach for targeting a single
program to multiple fundamentally different low level execution and
programming models while allowing the application programmer to adopt
a single high level programming model.  We suggest that features of
Fortran 90 for data parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
low-level programming models such as OpenCL and CUDA.

@@

A data parallel programming model is well suited to emerging architectures
that are based on either vector processing or massively parallel collections
of simple cores.  A popular trend for programming systems such as GPU
accelerators is the CUDA language from NVidia, or new languages like Chapel
from Cray.  A less intrusive approach that is well suited to legacy 
applications and languages are language extensions or libraries that allow
programmers to avoid adopting entirely new languages.  The recent OpenCL
specification is intended to support this programming model, as are
directive-based methods such as OpenMP or the ``Accelerate'' product from
the Portland Group.

@@

@@

\cite{chamberlain04zpl,pgi10accelerator}

>>>>

\title{Exploiting First-Class Arrays in Fortran for Accelerator Programming}

%%
%% reorder appropriately later
%%
\author{\IEEEauthorblockN{Matthew J. Sottile}
\IEEEauthorblockA{Galois, Inc.\\
421 SW 6th Ave. Suite 300 \\
Portland, OR 97204\\
Email: matt@galois.com}
\and
\IEEEauthorblockN{Craig E Rasmussen}
\IEEEauthorblockA{Los Alamos National Laboratory\\
CCS-7, MS B287\\
Los Alamos, NM\\
Email: rasmussn@lanl.gov}
}

@@

\begin{abstract}
Emerging architectures for high performance computing often are well
suited to a data parallel programming model.  This paper presents a
simple programming methodology based on existing languages and compiler
tools that allows programmers to take advantage of these systems.
We will work with the array features of Fortran 90 to show how this
infrequently exploited, standardized language feature is easily
transformed to lower level accelerator code.  Our transformations are
based on a mapping from Fortran 90 to C++ code with OpenCL extensions.
\end{abstract}

@@

This paper presents a compiler-level approach for targeting a single
program to multiple fundamentally different low level execution and
programming models while allowing the application programmer to adopt
a single high level programming model.  We show that features of
Fortran 90 for data parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
low-level programming models such as OpenCL and CUDA.  For algorithms
that can be easily expressed in terms of whole array, data parallel operations,
writing code in Fortran and transforming it automatically to specific low
level implementations removes the burden from the programmer of working
with tedious, error prone, low-level tools.

@@

A data parallel programming model is well suited to emerging
architectures that are based on either vector processing or massively
parallel collections of simple cores.  A popular trend for programming
systems such as GPU accelerators is the CUDA language from NVidia, or
new languages like Chapel from Cray.  A less intrusive approach that
is well suited to legacy applications and languages are language
extensions or libraries that allow programmers to avoid adopting
entirely new languages.  The recent OpenCL specification is intended
to support this programming model, as are directive-based methods such
as OpenMP or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}.

@@

Techniques based on whole-array programming being mapped to high performance,
parallel implementations are not new.

@@

\cite{chamberlain04zpl, roth97stencils}
